Quantitative Benchmarking for Automatically Generated Conjectures

### 2017-09-13 TEBenchmarkPaper/article.tex
More TEBenchmarks fleshing out

Existing attempts at tackling this problem have been found difficult to compare
and study, due partially to the variety of approaches taken, but also because of
the inherent ambiguity of the task and the different goals emphasised by their
authors and evaluation methods.

@@

\begin{itemize}
\item A general methodology for benchmarking ATE systems.
\item Resolving the issue of ``interestingness'' through the use theorem-proving
  benchmarks as a ground-truth.
\item A specific instantiation of this methodology, using the Tons of Inductive
  Problems benchmark as a corpus.
\item Automated tooling to produce this benchmark and analyse its output.
\item Application of this methodology to the QuickSpec and IsaCoSy ATE systems,
  and a discussion of the results.
\end{itemize}

@@

If we have a logical theory, for example the theory of lists shown in figure
\ref{figure:list_theory}, we may want to find theorems which describe its
behaviour. For a software library we may want to verify that certain
(un)desirable properties do (not) hold; we may also want to \emph{optimise}
programs using this library, rewriting expressions into a form which requires
less time, memory, network usage, etc. To avoid altering the program's result,
it is important to prove particular rewrite rules correct. Consider the
following theorem:

@@

This justifies a rewrite rule for splitting a single $\text{map}$ call into
multiple independent calls dealing with different sections of the list. This is
useful for the ``map/reduce'' programming paradigm, as each call can be
evaluated by a separate machine in parallel.

@@

Discovering theorems from an arbitrary theory requires two steps:

@@

Lots of research has been devoted to the second step, resulting in the field of
Automated Theorem Proving (ATP) and its spinoff applications. Less attention has
been paid to the first step, that of Automated Theory Exploration (ATE).

A major challenge in ATE is the requirement for conjectures to be
``interesting'', since this is an imprecise term with many different
interpretations, which is reflected in the variety of disparate approaches taken
so far.

A survey of these systems and their notions of ``interestingness'' is given by
Colton et al~\cite{colton2000notion}, for both conjectures and \emph{concepts}
(the contents of a theory, which for ATE we consider to be provided as input,
rather than part of the generating process). We summarise those aspects relevant
to conjecture formation in table \ref{table:colton}, along with later systems
we have studied.

@@

One aspect mentioned in the survey is that ``interestingness'' is not just a
measure of output quality, but can also impact the core design of a system. For
example, the use of constraint solving in IsaCoSy is to prevent forming
conjectures which are simple substitutions of previous ones; this is because the
designers consider such conjectures to be uninteresting. This further
complicates comparison of systems based on standalone evaluations, since
assumptions about such behaviour may be implicit in the interpretation (e.g. an
analysis of IsaCoSy conjectures does not need to consider substitutions), and
each methodology will be using different assumptions.

@@

The variety of existing ATE systems, their different domains, and their varying
interpretations of what is interesting makes it difficult to directly compare
their performance.

Whilst all offer insights into the ATE problem, we will limit ourselves to those
operating in higher-order logic and inductive theories, since this domain is the
most relevant for our interest in analysing software. The relevant systems in
this domain are QuickSpec, operating on functions in the Haskell programming
language; and IsaCoSy and IsaScheme, which operate in the Isabelle proof
assistant.

>>>>

Existing attempts at tackling this problem are difficult to compare, due
partially to the variety of approaches taken, but also because of the inherent
ambiguity of the task and the different goals emphasised by their designers and
their choice of evaluation method.

@@

\begin{itemize}
\item A general methodology for benchmarking ATE systems.
\item Resolving the issue of ``interestingness'' through the use of
  theorem-proving benchmarks as a ground-truth.
\item A specific instantiation of this methodology, using the Tons of Inductive
  Problems benchmark as a corpus.
\item Automated tooling to perform this benchmarking.
\item Application of our methodology to the QuickSpec and IsaCoSy ATE systems,
  and a discussion of the results.
\end{itemize}

@@

Given a logical theory, like the theory of lists shown in figure
\ref{figure:list_theory}, we may want to find theorems which describe its
behaviour. This could be for mathematical curiosity, or due to the theory's
importance in some domain. In particular, for theories which capture the
semantics of some software library, we may want to verify that certain
(un)desirable properties do (not) hold; we might also want to \emph{optimise}
programs using this library, rewriting expressions into a form which requires
less time, memory, network usage, etc. To avoid altering a program's result,
such rewrites should come with theorems proving their correctness, such as the
following theorem for our theory of lists:

@@

This justifies a rewrite rule for splitting a single $\text{map}$ call into
multiple independent calls dealing with different sections of a list. Such rules
are useful optimisations since each call can be evaluated in parallel, leading
to the ``map/reduce'' programming paradigm.

All of these example use cases require the ability to discover theorems about
some particular theory. This is a hard problem in general, but presents
opportunities for automation due to the precise, symbolic nature of the domain.

@@

The task of discovering theorems in an arbitrary theory can be broken down into
two steps:

@@

This second step is studied extensively in the field of Automated Theorem
Proving (ATP). Less attention has been paid to the first step, that of
Automated Theory Exploration (ATE).

A major challenge in ATE is choosing how to narrow down the set of generated
conjectures to those deemed ``interesting'', since this is an imprecise term
with many different interpretations. For example, all existing approaches agree
that simple tautologies are ``uninteresting'', but differ when it comes to more
complex statements.

A survey of ATE systems and their notions of ``interestingness'' is given by
Colton et al~\cite{colton2000notion}. Their survey also includes \emph{concept
  formation}, which corresponds to the automatic creation of theories, although
We ignore this as a separate concern: the ATE task assumes that a theory is
provided. We summarise those aspects relevant to conjecture formation in table
\ref{table:colton}, and extend the comparison with three more recent systems.

@@

@@

The criteria for ``interestingness'' in table \ref{table:colton} not only allow
the results of ATE systems to be evaluated, but they also affect the choice of
algorithm used by those systems in the first place. This coupling between
implementation detail and assessment method makes it difficult to compare
existing approaches, since the assumptions made by each evaluation methodology
may lead to incompatible results. For example, the design of IsaCoSy treats
simple substitutions of existing conjectures as uninteresting; yet this
criterion is absent from its evaluation, since the algorithm uses a constraint
solver to entirely avoid such results.

If we limit ourselves to exploring higher-order theories with inductive types
(a domain closely matching our interest in analysing software), we do find a
direct comparison of three systems~\cite{claessen2013automating}: QuickSpec (the
conjecture-generating component of HipSpec), IsaCoSy and IsaScheme. This is a
\emph{precision/recall} analysis, providing the same theories to each system and
comparing the set of generated conjectures against a \emph{ground truth}.
